---
title: "Mismatch Index: When Environment, Health, and Wealth Don't Align"
author: "Mini Project 2"
date: "`r Sys.Date()`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    toc: true
    theme:
      version: 4
      bootswatch: flatly
      primary: "#ed4a1a"
      secondary: "#48bea1"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{css}
/* I am putting this chunk here in case I need to create a functional knitr::kable() table */
.chart-wrapper {
  overflow-y: auto;
}

/* set a custom font for banner */
.navbar { font-family: "Helvetica Neue", Helvetica, Arial, sans-serif; }
```

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load packages
library(tidyverse)
library(broom)
library(plotly)
library(flexdashboard)
library(DT)
library(sf)
library(ggplot2)
library(dplyr)
library(ggcorrplot)
library(patchwork)

# load data
url <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/health_data.csv"
health_data <- read.csv(url)

# read shapefile of Chicago
chicago_sf <- st_read("https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/chi_comm_areas.geojson", quiet = TRUE)

# join health data with shapefile
chicago_map <- chicago_sf %>%
  left_join(health_data %>%
              mutate(GEOID = as.character(GEOID)), by = c("area_numbe" = "GEOID"))

# pull hardship index from most recent census data from github repository
url2 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/h_index.csv"
h_index <- read.csv(url2) %>%
  slice(-1, ) # remove first row

# join hardship index to health + map data
chicago_map <- chicago_map %>%
  left_join(h_index %>% select(GEOID, `HDX_2020.2024`) %>% mutate(GEOID = as.character(GEOID), `HDX_2020.2024` = as.numeric(`HDX_2020.2024`)), by = c("area_numbe" = "GEOID"))

# transform data into percents
chicago_map <- chicago_map %>%
  mutate(
    diabetes_rate = HCSDIA_2023.2024 / Population * 100,
    obesity_rate = HCSOB_2023.2024 / Population * 100,
    hypertension_rate = HCSHYT_2023.2024 / Population * 100,
    asthma_rate = HCSATH_2023.2024 / Population * 100,
    hard_index = HDX_2020.2024,
    pct_white = PCT.W_2020.2024,
    disease_burden = rowMeans(cbind(diabetes_rate, obesity_rate, hypertension_rate, asthma_rate), na.rm = TRUE)
  ) %>% # add resilience and vulnerability scores
  mutate(
    resilient = hard_index > quantile(hard_index, 0.5, na.rm = TRUE) &
                disease_burden < quantile(disease_burden, 0.5, na.rm = TRUE),
    vulnerable = hard_index < quantile(hard_index, 0.5, na.rm = TRUE) &
                 disease_burden > quantile(disease_burden, 0.5, na.rm = TRUE)
  ) %>% # add diabetes specific resilience and vulnerability scores
  mutate(
    resilient_dia = hard_index > quantile(hard_index, 0.5, na.rm = TRUE) &
                    diabetes_rate < quantile(diabetes_rate, 0.5, na.rm = TRUE),
    vulnerable_dia = hard_index < quantile(hard_index, 0.5, na.rm = TRUE) &
                     diabetes_rate > quantile(diabetes_rate, 0.5, na.rm = TRUE),
    category_dia = case_when(
      resilient_dia == TRUE ~ "Resilient",
      vulnerable_dia == TRUE ~ "Vulnerable",
      TRUE ~ "Expected"
    )
  )

# remove Fuller Park and Burnside - raw number data do not match CHI Atlas rates, suspected data quality issue
chicago_map <- chicago_map %>%
  filter(!community %in% c("FULLER PARK", "BURNSIDE"))

# set up cor data
cor_data <- chicago_map %>%
  st_drop_geometry() %>%
  select(
    Diabetes = diabetes_rate,
    Obesity = obesity_rate,
    Hypertension = hypertension_rate,
    Asthma = asthma_rate,
    Pct_White = PCT.W_2020.2024,
    Air_Quality = CHASBQJ_2023.2024,
    Traffic_Risk = TRF_2020,
    Env_Justice = CHAKNKC_2023,
    Hardship = hard_index,
    Community = Name
  ) %>%
  drop_na()

# make custom color palettes
palette1 <- c("#2e4f4f", "#0f6340", "#3f6d12", "#6f6e00", "#976000", "#c95500", "#ff0000")

palette2 <- c("#2f4f4f", "#736e00", "#ed4a1a", "#ff7e5d", "#ffb198", "#ffdacd", "#ffffff")

palette3 <- c("#2d4f4f", "#ef0300", "#48bea1", "#167b2b", "#a7e831", "#7b4419", "#f4bb8f")
```

## Row {data-height="400"}

### Disease Burden by Community Area

```{r disease maps, echo=FALSE, out.width="100%", results='asis'}
# create chloropleth function
make_map <- function(var, title, midpt) {
  ggplot(chicago_map) +
    geom_sf(aes(fill = .data[[var]],
                text = paste0("<b>", Name, "</b><br>",
                              title, ": ", round(.data[[var]], 1), "%"))) +
    scale_fill_gradient2(low = "#2e4f4f",
                         mid = "#f4bb8f",
                         high = "#ff0000",
                         midpoint = midpt,
                         na.value = "black",
                         name = "% Adults") +
    theme_void() +
    theme(legend.position = "none")
}

# construct maps
pd <- make_map("diabetes_rate", "Diabetes", mean(chicago_map$diabetes_rate, na.rm = TRUE))
po <- make_map("obesity_rate", "Obesity", mean(chicago_map$obesity_rate, na.rm = TRUE))
ph <- make_map("hypertension_rate", "Hypertension", mean(chicago_map$hypertension_rate, na.rm = TRUE))
pa <- make_map("asthma_rate", "Asthma", mean(chicago_map$asthma_rate, na.rm = TRUE))

# generate interactive maps of spatial patterns of disease
pd2 <- ggplotly(pd, tooltip = "text") %>% layout(autosize = TRUE)
po2 <- ggplotly(po, tooltip = "text") %>% layout(autosize = TRUE)
ph2 <- ggplotly(ph, tooltip = "text") %>% layout(autosize = TRUE)
pa2 <- ggplotly(pa, tooltip = "text") %>% layout(autosize = TRUE)

# arrange plots in a line
subplot(pd2, po2, ph2, pa2, nrows = 1) %>%
  layout(
    autosize = TRUE,
    annotations = list(
      list(x = 0.125, y = 1.0, text = "<b>Diabetes</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.375, y = 1.0, text = "<b>Obesity</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.625, y = 1.0, text = "<b>Hypertension</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419")),
      list(x = 0.875, y = 1.0, text = "<b>Asthma</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center",
           font = list(size = 14, color = "#7b4419"))
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Mean Diabetes Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$diabetes_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Diabetes Rate",
  icon = "fa-droplet",
  color = "#2d4f4f"
)
```

### Mean Obesity Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$obesity_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Obesity Rate",
  icon = "fa-weight-hanging",
  color = "#f4bb8f"
)
```

### Mean Hypertension Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$hypertension_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Hypertension Rate",
  icon = "fa-heart-pulse",
  color = "#2d4f4f"
)
```

### Mean Asthma Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$asthma_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Asthma Rate",
  icon = "fa-lungs",
  color = "#f4bb8f"
)
```

## Row {data-height="600"}

### Summary Table of Disease Burden

```{r top diabetes neighborhoods}
chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(diabetes_rate)) %>%
  arrange(desc(diabetes_rate)) %>%
  select(Community = community, Diabetes = diabetes_rate, Obesity = obesity_rate, Hypertension = hypertension_rate, Asthma = asthma_rate) %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  DT::datatable(options = list(
  pageLength = 67,
  dom = 't',
  scrollY = "450px",
  scrollCollapse = TRUE
))
```

### Spatial Determinants of Disease Burden 

``` {r chloropleth disease burden }
# calculate mean of all disease rates
chicago_map <- chicago_map %>%
  mutate(disease_burden = rowMeans(cbind(diabetes_rate, obesity_rate, hypertension_rate, asthma_rate), na.rm = TRUE))

# visualize on map
disease_burden <- ggplot(chicago_map) +
  geom_sf(aes(fill = disease_burden,
              text = paste0("<b>", Name, "</b><br>",
              "<b>Disease Burden:</b> ", round(disease_burden, 1), "%"))) +
  scale_fill_gradient2(low = "#2e4f4f", mid = "#f4bb8f", high = "#ef0300",
                       midpoint = mean(chicago_map$disease_burden, na.rm = TRUE),
                       na.value = "black", name = NA) +
  theme_void() +
  theme(plot.title = element_text(size = 14, color = "#7b4419", face = "bold", hjust = 0.5)) +
  labs(title = "Overall Disease Burden by Community Area")

ggplotly(disease_burden, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Highest Overall Burden

```{r}
top_burden <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(disease_burden)) %>%
  slice_max(disease_burden, n = 1)

valueBox(
  value = paste0(round(top_burden$disease_burden, 1), "%"),
  caption = paste0("Highest Overall Burden — ", top_burden$community),
  icon = "fa-location-dot",
  color = "#7b4419"
)
```

### Lowest Overall Burden

```{r}
bot_burden <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(disease_burden)) %>%
  slice_min(disease_burden, n = 1)

valueBox(
  value = paste0(round(bot_burden$disease_burden, 1), "%"),
  caption = paste0("Lowest Overall Burden — ", bot_burden$community),
  icon = "fa-location-dot",
  color = "#48bea1"
)
```

### Highest Hardship Index

```{r}
top_hardship <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(hard_index)) %>%
  slice_max(hard_index, n = 1)

valueBox(
  value = round(top_hardship$hard_index, 1),
  caption = paste0("Highest Hardship — ", top_hardship$community),
  icon = "fa-location-dot",
  color = "#ed4a1a"
)
```

### Lowest Hardship Index

```{r}
bot_hardship <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(hard_index)) %>%
  slice_min(hard_index, n = 1)

valueBox(
  value = round(bot_hardship$hard_index, 1),
  caption = paste0("Lowest Hardship — ", bot_hardship$community),
  icon = "fa-location-dot",
  color = "#167b2b"
)
```
## Row {data-height="450"}

### Racial Determinants of Disease Burden and Hardship

``` {r scatterplot disease burden vs hardship with percent white}
scatter <- ggplot(chicago_map, aes(x = hard_index, y = disease_burden, 
                                    color = pct_white,
                                    text = paste0("<b>", community, "</b><br>",
                                                  "Hardship: ", round(hard_index, 1), "<br>",
                                                  "Disease Burden: ", round(disease_burden, 1), "%<br>",
                                                  "% White: ", round(PCT.W_2020.2024, 1), "%"))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", color = "black") +
  scale_color_gradient(low = "#753D12", high = "#ffffff", name = "% White") +
  labs(x = "Hardship Index", y = "Disease Burden (%)") +
  theme_minimal()

ggplotly(scatter, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Correlation Matrix

```{r corr plot}
# create a correlation matrix of all variables
cor_matrix <- cor(cor_data %>%
                    select(-Community)
                  )

p_cor <- ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray"),
           title = "Correlation Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Correlation: % White & Disease Burden

```{r}
cor_val <- cor(chicago_map$PCT.W_2020.2024, chicago_map$disease_burden, use = "complete.obs")

valueBox(
  value = round(cor_val, 2),
  caption = "Correlation: % White & Disease Burden",
  icon = "fa-chart-line",
  color = "#2d4f4f"
)
```

### Correlation: % White & Hardship

```{r}
cor_val2 <- cor(chicago_map$PCT.W_2020.2024, chicago_map$hard_index, use = "complete.obs")

valueBox(
  value = round(cor_val2, 2),
  caption = "Correlation: % White & Hardship Index",
  icon = "fa-chart-line",
  color = "#ed4a1a"
)
```

### Correlation: Hardship & Disease Burden

```{r}
cor_val3 <- cor(chicago_map$hard_index, chicago_map$disease_burden, use = "complete.obs")

valueBox(
  value = round(cor_val3, 2),
  caption = "Correlation: Hardship & Disease Burden",
  icon = "fa-chart-line",
  color = "#f4bb8f"
)
```

### Number of Neighborhoods Above Trend

```{r}
model_burden <- lm(disease_burden ~ hard_index, data = chicago_map)
above <- sum(residuals(model_burden) > 0, na.rm = TRUE)

valueBox(
  value = above,
  caption = "Neighborhoods Above Expected Burden",
  icon = "fa-triangle-exclamation",
  color = "#7b4419"
)
```

## Row {data-height="600"}

### Intra-Categorical Correlation Plot: Disease

``` {r intracat corr disease plot}
# corrplot disease
cor_matrix_disease <- cor(cor_data %>%
                            select(Diabetes, Obesity, Hypertension, Asthma)
                          )

#plot with ggcorrplot
p_cor_disease <- ggcorrplot(cor_matrix_disease,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray"),
           title = "Intracategorical Correlation Among Diseases") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# ggplotly it
ggplotly(p_cor_disease) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```

### Disease vs. Environmental Factors

``` {r disease v environment}

# create vectors of two categories
sub_disease <- c("Diabetes", "Obesity", "Hypertension", "Asthma")
sub_environ <- c("Pct_White", "Air_Quality", "Traffic_Risk", "Env_Justice", "Hardship")

cor_matrix2 <- cor(cor_data %>%
                    select(all_of(c(sub_disease, sub_environ))),
                  use = "complete.obs")

p_cor_matrix2 <- ggcorrplot(cor_matrix2[sub_disease, sub_environ],
           colors = c("darkslategray", "white", "red2")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_line(color = "white", linewidth = 1))

ggplotly(p_cor_matrix2) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Intra-Categorical Correlation Plot: Social Determinants of Health

``` {r intra corr environmental factors plot}
cor_matrix_sdoh <- cor(cor_data %>%
                            select(Pct_White, Air_Quality, Traffic_Risk, Env_Justice, Hardship)
                          )

p_cor_sdoh <- ggcorrplot(cor_matrix_sdoh,
           method = "circle",
           type = "lower",
           lab = FALSE,
           lab_size = 3,
           colors = c("red2", "white", "darkslategray"),
           title = "Intracategorical Correlation Among Social Determinants of Health") +
  scale_x_discrete(limits = rev) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor_sdoh) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```

## Row {data-height="450"}

### Diabetes Mismatch Index

``` {r diabetes residual map and coef plot}
# Fit a linear regression model diabetes ~ hardship
dia_model <- lm(diabetes_rate ~ hard_index, data = chicago_map)
# extract model
dia_tbl <- tidy(dia_model, conf.int = TRUE)
# add residuals to df
chicago_map$residual[as.numeric(rownames(dia_model$model))] <- residuals(dia_model)
# Remove intercept and plot pointranges (estimate + CI) with a vertical line at 0
dia_tbl <- dia_tbl %>%
  filter(term != "(Intercept)")

# ggplot of regression
dia_hard <- ggplot(dia_tbl, aes(x = term, y = estimate)) +
  geom_pointrange(aes( ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "#e03000") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Regression Coefficient with 95% CI",
       y = "Estimate",
       x = "Term")

# plot interactive
ggplotly(dia_hard) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Diabetes Mismatch Index

``` {r diabetes residuals}
# residual tooltip
chicago_map <- chicago_map %>%
  filter(!is.na(diabetes_rate)) %>%
  mutate(
    tooltip_text = paste0(
      "<b>", community, "</b><br>",
      "Diabetes Rate: ", round(diabetes_rate, 1), "%<br>",
      "Predicted: ", round(diabetes_rate - residual, 1), "%<br>",
      "Residual: ", round(residual, 1)
    )
  )

diabetes_resid <- ggplot(chicago_map) +
  geom_sf(aes(fill = residual, text = tooltip_text)) +
  scale_fill_gradient2(
    low = "#48bea1", mid = "white", high = "#7b4419",
    midpoint = 0, name = "Residual"
  ) +
  theme_minimal() +
  labs(title = "Diabetes Mismatch Index",
       subtitle = "Dark gray = worse than expected | Red = better than expected")

ggplotly(diabetes_resid, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Resilient Neighborhoods

```{r}
n_resilient <- sum(chicago_map$resilient_dia == TRUE, na.rm = TRUE)
valueBox(
  value = n_resilient,
  caption = "Resilient Neighborhoods",
  icon = "fa-shield",
  color = "#48bea1"
)
```

### Vulnerable Neighborhoods

```{r}
n_vulnerable <- sum(chicago_map$vulnerable_dia == TRUE, na.rm = TRUE)
valueBox(
  value = n_vulnerable,
  caption = "Vulnerable Neighborhoods",
  icon = "fa-triangle-exclamation",
  color = "#7b4419"
)
```

### Neighborhoods Outside Expected Pattern

```{r}
n_mismatch <- sum(chicago_map$resilient_dia == TRUE | chicago_map$vulnerable_dia == TRUE, na.rm = TRUE)
valueBox(
  value = n_mismatch,
  caption = "Neighborhoods Outside Expected Pattern",
  icon = "fa-shuffle",
  color = "#a7e831"
)
```

### Variance in Diabetes Explained by Hardship

```{r}
valueBox(
  value = "39%",
  caption = "Variance in Diabetes Explained by Hardship",
  icon = "fa-chart-line",
  color = "#167b2b"
)
```

## Row {data-height="450"}

### Neighborhoods of Vulnerability and Resilience

```{r mismatch scatterplot}
scatter_mismatch <- ggplot(chicago_map, aes(x = hard_index, y = disease_burden,
                                             color = category_dia,
                                             text = paste0("<b>", community, "</b><br>",
                                                           "Hardship: ", round(hard_index, 1), "<br>",
                                                           "Disease Burden: ", round(disease_burden, 1), "%<br>",
                                                           category_dia))) +
  geom_point(size = 3) +
  geom_vline(xintercept = median(chicago_map$hard_index, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = median(chicago_map$disease_burden, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Resilient" = "#48bea1", 
                                "Vulnerable" = "#D10300", 
                                "Expected" = "grey70")) +
  labs(x = "Hardship Index", y = "Disease Burden (%)", color = NULL) +
  theme_minimal()

ggplotly(scatter_mismatch, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

### Diabetes-Specific Vulnerability and Resilience

``` {r scatterplot of diabetes vs hardship}
scatter_diabetes <- ggplot(chicago_map, aes(x = hard_index, y = diabetes_rate,
                                             color = category_dia,
                                             text = paste0("<b>", community, "</b><br>",
                                                           "Hardship: ", round(hard_index, 1), "<br>",
                                                           "Diabetes Rate: ", round(diabetes_rate, 1), "%<br>",
                                                           category_dia))) +
  geom_point(size = 3) +
  geom_vline(xintercept = median(chicago_map$hard_index, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = median(chicago_map$diabetes_rate, na.rm = TRUE), 
             linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Resilient" = "#48bea1", 
                                "Vulnerable" = "#7b4419", 
                                "Expected" = "grey70")) +
  scale_x_discrete(limits = rev) +
  labs(x = "Hardship Index", y = "Diabetes Rate (%)", color = NULL) +
  theme_minimal()

ggplotly(scatter_diabetes, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```
