---
title: "Mismatch Index: When Environment, Health, and Wealth Don't Align"
author: "Mini Project 2"
date: "`r Sys.Date()`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    toc: true
    theme:
      version: 4
      bootswatch: flatly
      primary: "#D10300"
      secondary: "#48bea1"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{css}
/* I am putting this chunk here in case I need to create a functional knitr::kable() table */
.chart-wrapper {
  overflow-y: auto;
}

/* set a custom font for banner */
.navbar { font-family: "Helvetica Neue", Helvetica, Arial, sans-serif; }
```

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load packages
library(tidyverse)
library(broom)
library(plotly)
library(flexdashboard)
library(DT)
library(sf)
library(ggplot2)
library(dplyr)
library(plotly)
library(ggcorrplot)
library(patchwork)

# load data
url <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/health_data.csv"
health_data <- read.csv(url)
chicago_sf <- st_read("https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/chi_comm_areas.geojson", quiet = TRUE)
#left-join map data to health data
chicago_map <- chicago_sf %>%
  left_join(health_data %>% mutate(GEOID = as.character(GEOID)), by = c("area_numbe" = "GEOID"))

# grab the hardship index from most recent census data from github repository
url2 <- "https://raw.githubusercontent.com/fitzley/miniproj2/refs/heads/main/h_index.csv"
h_index <- read.csv(url2) %>%
  slice(-1, )
# left join hardship index from most recent census data
chicago_map <- chicago_map %>%
  left_join(h_index %>% select(GEOID, `HDX_2020.2024`) %>% mutate(GEOID = as.character(GEOID), `HDX_2020.2024` = as.numeric(`HDX_2020.2024`)), by = c("area_numbe" = "GEOID"))

# translate count data into rates for each disease
chicago_map <- chicago_map %>%
  mutate(
    diabetes_rate = HCSDIA_2023.2024 / Population * 100,
    obesity_rate = HCSOB_2023.2024 / Population * 100,
    hypertension_rate = HCSHYT_2023.2024 / Population * 100,
    asthma_rate = HCSATH_2023.2024 / Population * 100,
    hard_index = HDX_2020.2024
  )

# remove Fuller Park and Burnside - raw number data do not match CHI Atlas rates, suspected data quality issue
chicago_map <- chicago_map %>%
  filter(!community %in% c("FULLER PARK", "BURNSIDE"))

# set up cor data
cor_data <- chicago_map %>%
  st_drop_geometry() %>%
  select(
    Diabetes = diabetes_rate,
    Obesity = obesity_rate,
    Hypertension = hypertension_rate,
    Asthma = asthma_rate,
    Pct_White = PCT.W_2020.2024,
    Air_Quality = CHASBQJ_2023.2024,
    Traffic_Risk = TRF_2020,
    Env_Justice = CHAKNKC_2023,
    Hardship = hard_index,
    Community = Name
  ) %>%
  drop_na()

# make custom color palettes
palette1 <- c("#2e4f4f", "#0f6340", "#3f6d12", "#6f6e00", "#976000", "#c95500", "#ff0000")
palette2 <- c("#2f4f4f", "#736e00", "#ed4a1a", "#ff7e5d", "#ffb198", "#ffdacd", "#ffffff")
palette3 <- c("#2d4f4f", "#ef0300", "#48bea1", "#167b2b", "#a7e831", "#7b4419", "#f4bb8f")
```

## Row {data-height="400"}

### Disease Burden by Community Area

```{r disease maps, echo=FALSE, out.width="100%", results='asis'}
# create chloropleth function
make_map <- function(var, title, midpt) {
  ggplot(chicago_map) +
    geom_sf(aes(fill = .data[[var]],
                text = paste0("<b>", community, "</b><br>",
                              title, ": ", round(.data[[var]], 1), "%"))) +
    scale_fill_gradient2(low = "#2e4f4f", mid = "white", high = "#ff0000", midpoint = midpt, na.value = "black", name = "% Adults") +
    theme_void() +
    theme(legend.position = "none")
}

# make maps
pd <- make_map("diabetes_rate", "Diabetes", mean(chicago_map$diabetes_rate, na.rm = TRUE))
po <- make_map("obesity_rate", "Obesity", mean(chicago_map$obesity_rate, na.rm = TRUE))
ph <- make_map("hypertension_rate", "Hypertension", mean(chicago_map$hypertension_rate, na.rm = TRUE))
pa <- make_map("asthma_rate", "Asthma", mean(chicago_map$asthma_rate, na.rm = TRUE))

# make maps interactive
pd2 <- ggplotly(pd, tooltip = "text") %>% layout(autosize = TRUE)
po2 <- ggplotly(po, tooltip = "text") %>% layout(autosize = TRUE)
ph2 <- ggplotly(ph, tooltip = "text") %>% layout(autosize = TRUE)
pa2 <- ggplotly(pa, tooltip = "text") %>% layout(autosize = TRUE)

# plot in a line
subplot(pd2, po2, ph2, pa2, nrows = 1) %>%
  layout(
    autosize = TRUE,
    annotations = list(
      list(x = 0.125, y = 1.05, text = "<b>Diabetes</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center"),
      list(x = 0.375, y = 1.05, text = "<b>Obesity</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center"),
      list(x = 0.625, y = 1.05, text = "<b>Hypertension</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center"),
      list(x = 0.875, y = 1.05, text = "<b>Asthma</b>",
           showarrow = FALSE, xref = "paper", yref = "paper", xanchor = "center")
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Mean Diabetes Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$diabetes_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Diabetes Rate",
  icon = "fa-droplet",
  color = "primary"
)
```

### Mean Obesity Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$obesity_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Obesity Rate",
  icon = "fa-weight-hanging",
  color = "primary"
)
```

### Mean Hypertension Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$hypertension_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Hypertension Rate",
  icon = "fa-heart-pulse",
  color = "primary"
)
```

### Mean Asthma Rate

```{r}
valueBox(
  value = paste0(round(mean(chicago_map$asthma_rate, na.rm = TRUE), 1), "%"),
  caption = "Mean Asthma Rate",
  icon = "fa-lungs",
  color = "primary"
)
```

## Row {data-height="600"}

### Summary Table of Disease Burden

```{r top diabetes neighborhoods}
chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(diabetes_rate)) %>%
  arrange(desc(diabetes_rate)) %>%
  select(Community = community, Diabetes = diabetes_rate, Obesity = obesity_rate, Hypertension = hypertension_rate, Asthma = asthma_rate) %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  DT::datatable(options = list(
  pageLength = 67,
  dom = 't',
  scrollY = "450px",
  scrollCollapse = TRUE
))
```

### Correlation Matrix

```{r corr plot}
# create a correlation matrix of all variables
cor_matrix <- cor(cor_data %>%
                    select(-Community)
                  )

p_cor <- ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray"),
           title = "Correlation Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="600"}

### Intra-Categorical Correlation Plot: Disease

``` {r intracat corr disease plot}
# corrplot disease
cor_matrix_disease <- cor(cor_data %>%
                            select(Diabetes, Obesity, Hypertension, Asthma)
                          )

#plot with ggcorrplot
p_cor_disease <- ggcorrplot(cor_matrix_disease,
           method = "circle",
           type = "lower",
           lab = FALSE,
           colors = c("red2", "white", "darkslategray"),
           title = "Intracategorical Correlation Among Diseases") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# ggplotly it
ggplotly(p_cor_disease) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```

### Intra-Categorical Correlation Plot: Social Determinants of Health

``` {r intra corr environmental factors plot}
cor_matrix_sdoh <- cor(cor_data %>%
                            select(Pct_White, Air_Quality, Traffic_Risk, Env_Justice, Hardship)
                          )

p_cor_sdoh <- ggcorrplot(cor_matrix_sdoh,
           method = "circle",
           type = "lower",
           lab = FALSE,
           lab_size = 3,
           colors = c("red2", "white", "darkslategray"),
           title = "Intracategorical Correlation Among Social Determinants of Health") +
  scale_x_discrete(limits = rev) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor_sdoh) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

```
## Row {data-height="450"}

### Disease vs. Environmental Factors

``` {r disease v environment}

# create vectors of two categories
sub_disease <- c("Diabetes", "Obesity", "Hypertension", "Asthma")
sub_environ <- c("Pct_White", "Air_Quality", "Traffic_Risk", "Env_Justice", "Hardship")

cor_matrix2 <- cor(cor_data %>%
                    select(all_of(c(sub_disease, sub_environ))),
                  use = "complete.obs")

p_cor_matrix2 <- ggcorrplot(cor_matrix2[sub_disease, sub_environ],
           colors = c("darkslategray", "white", "red2")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_cor_matrix2) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Highest Diabetes Burden

```{r}
top_diabetes <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(diabetes_rate)) %>%
  slice_max(diabetes_rate, n = 1)

valueBox(
  value = paste0(round(top_diabetes$diabetes_rate, 1), "%"),
  caption = paste0("Highest Diabetes — ", top_diabetes$community),
  icon = "fa-location-dot",
  color = "danger"
)
```

### Highest Obesity Burden

```{r}
top_obesity <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(obesity_rate)) %>%
  slice_max(obesity_rate, n = 1)

valueBox(
  value = paste0(round(top_obesity$obesity_rate, 1), "%"),
  caption = paste0("Highest Obesity — ", top_obesity$community),
  icon = "fa-location-dot",
  color = "danger"
)
```

### Highest Hypertension Burden

```{r}
top_hypertension <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(hypertension_rate)) %>%
  slice_max(hypertension_rate, n = 1)

valueBox(
  value = paste0(round(top_hypertension$hypertension_rate, 1), "%"),
  caption = paste0("Highest Hypertension — ", top_hypertension$community),
  icon = "fa-location-dot",
  color = "danger"
)
```

### Highest Asthma Burden

```{r}
top_asthma <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(asthma_rate)) %>%
  slice_max(asthma_rate, n = 1)

valueBox(
  value = paste0(round(top_asthma$asthma_rate, 1), "%"),
  caption = paste0("Highest Asthma — ", top_asthma$community),
  icon = "fa-location-dot",
  color = "danger"
)
```

## Row

### Diabetes Mismatch Index

```{r diabetes mismatch, echo=FALSE}

# Fit a multiple regression model
dia_model_big <- lm(Diabetes ~ Hardship + Air_Quality, data = cor_data)
big_tbl <- tidy(dia_model_big, conf.int = TRUE)

# Remove intercept
# Plot pointranges (estimate + CI) with a vertical line at 0
big_tbl %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_pointrange(aes( ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "#ee0000") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Multiple Regression Coefficients with 95% CI",
       y = "Estimate",
       x = "Term")

# # plot Diabetes vs % white
#  ggplot(cor_data, aes(x = Diabetes, y = Pct_White)) +
#    geom_point(alpha = 0.7) +
#    geom_smooth(method = lm) +
#    theme_minimal() +
#    labs(x = "Adult Diabetes Rate", y = "% White Individuals in Community")

# Fit a linear regression model diabetes ~ hardship
dia_model <- lm(diabetes_rate ~ hard_index, data = chicago_map)
# extract model
dia_tbl <- tidy(dia_model, conf.int = TRUE)
# add residuals to df
chicago_map$residual[as.numeric(rownames(dia_model$model))] <- residuals(dia_model)
# Remove intercept and plot pointranges (estimate + CI) with a vertical line at 0
dia_tbl %>%
  filter(term != "(Intercept)")

# ggplot of regression
dia_hard <- ggplot(dia_tbl, aes(x = term, y = estimate)) +
  geom_pointrange(aes( ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "#ee0000") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Regression Coefficient with 95% CI",
       y = "Estimate",
       x = "Term")

# plot interactive
ggplotly(dia_hard) %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)

# residual tooltip
chicago_map <- chicago_map %>%
  filter(!is.na(diabetes_rate)) %>%
  mutate(
    tooltip_text = paste0(
      "<b>", community, "</b><br>",
      "Diabetes Rate: ", round(diabetes_rate, 1), "%<br>",
      "Predicted: ", round(diabetes_rate - residual, 1), "%<br>",
      "Residual: ", round(residual, 1)
    )
  )

diabetes_resid <- ggplot(chicago_map) +
  geom_sf(aes(fill = residual, text = tooltip_text)) +
  scale_fill_gradient2(
    low = "#48bea1", mid = "white", high = "#7b4419",
    midpoint = 0, name = "Residual"
  ) +
  theme_minimal() +
  labs(title = "Diabetes Mismatch Index",
       subtitle = "Dark gray = worse than expected | Red = better than expected")

ggplotly(diabetes_resid, tooltip = "text") %>%
  layout(autosize = TRUE) %>%
  config(displayModeBar = FALSE)
```

## Row {data-height="120"}

### Community Areas Analyzed

```{r}
n_areas <- sum(!is.na(chicago_map$diabetes_rate))
valueBox(
  value = n_areas,
  caption = "Community Areas Analyzed",
  icon = "fa-map-marker",
  color = "info"
)
```

### Mean Diabetes Rate

```{r}
mean_rate <- round(mean(chicago_map$diabetes_rate, na.rm = TRUE), 1)
valueBox(
  value = paste0(mean_rate, "%"),
  caption = "Mean Diabetes Rate",
  icon = "fa-heartbeat",
  color = "primary"
)
```

### Most Vulnerable Neighborhood

```{r}
most_vulnerable <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(residual), !is.na(diabetes_rate)) %>%
  slice_max(residual, n = 1) %>%
  pull(community)
valueBox(
  value = most_vulnerable,
  caption = "Highest Mismatch (Vulnerable)",
  icon = "fa-arrow-up",
  color = "danger"
)
```

### Most Resilient Neighborhood

```{r}
most_resilient <- chicago_map %>%
  st_drop_geometry() %>%
  filter(!is.na(residual), !is.na(diabetes_rate)) %>%
  slice_min(residual, n = 1) %>%
  pull(community)
valueBox(
  value = most_resilient,
  caption = "Highest Mismatch (Resilient)",
  icon = "fa-arrow-down",
  color = "success"
)
```

